---
title: "PCA"
---

A certain system can be in a continuum of different internal states or configurations. The set of all possible states can be considered as a manifold or a probability distribution in a low-dimensional space $\mathbb{R}^d$. From outside, the system's response to situation or features of its outwards behaviour can be studied in a multitude of way, yielding scores on a large number of possible assesment procedures, each 

https://projects.ori.org/lrg/PDFs_papers/Goldberg.Big-Five-Markers-Psych.Assess.1992.pdf 
(100 markers in Appendix A)

Example: Personality:
- external view: subjects' answers to a large number of questions
- internal view: a smaller number of personality dimensions

### Abstract simulated example

We have 1000 objects, whose internal state can be described by a 3D vector

```{r}
n <- 1000
true_states <- cbind(
  rnorm( n ),
  ifelse( runif(n)<.3, rnorm( n, -1, 2 ), rnorm( n, 3, .5 ) ),
  runif( n, -2, 2 )
)

hist(x[,3])
```

We have 7 different possible measurements whose results depend in different ways on the state:
```{r}
true_loadings <- rbind(
  c( 1, .5, 0 ),
  c( -20, 0, 10 ),
  c( -1, .1, .1 ),
  c( 0, 9, 2 ),
  c( 1, 1, 1 ),
  c( .1, -.1, 3 ),
  c( 0, 0, .8 )
)  

measurement_noise_sd <- .1 * c( .1, .2, .5, .3, .4, .15, .4 ) 

y <- true_states %*% t(true_loadings) + rnorm( n*7, 0, rep( measurement_noise_sd, each=n ) )
```

```{r}
pca <- prcomp( y, center=TRUE, scale. = TRUE )
pca
```

```{r}
cor(true_states, pca$x )
```

```{r}
true_states <- matrix( rnorm( 3*n ), ncol=3 )
true_loadings <- matrix( rnorm( 3*7 ), ncol=3 )
y <- true_states %*% t(true_loadings)
pca <- prcomp( y ) 
pca
```

